---
title: "Prediction Assignment Writeup"
author: "frsmeireles"
date: "9/12/2020"
output: html_document
---

## Overview


The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).


## Data

The training data for this project are available at:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available at:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


## Getting and Cleaning Data

The first step was to download the data, load it into R and prepare it for the modeling process.

**1 - Load data**

Data, all functions and packages are loaded. 

```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(e1071)
library(randomForest)
set.seed(1)

train.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test.url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

train.data.raw <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!",""))
test.data.raw <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!",""))
```

**2 - Remove unecessary colums**

Columns that are not needed for the model and columns that contain NAs ahould be removed.

```{r}
train.clean1 <- train.data.raw[,8:length(colnames(train.data.raw))]
test.clean1 <- test.data.raw[,8:length(colnames(test.data.raw))]

train.clean1 <- train.clean1[, colSums(is.na(train.clean1)) == 0] 
test.clean1 <- test.clean1[, colSums(is.na(test.clean1)) == 0] 

nzv <- nearZeroVar(train.clean1,saveMetrics=TRUE)
zero.var.ind <- sum(nzv$nzv)

if ((zero.var.ind>0)) {
        train.clean1 <- train.clean1[,nzv$nzv==FALSE]
}
```

**3 - Slice the data for cross validation**

After that, the next step is to divide the data into two sets: training and validation.
This first is a training set with 70% of the data which is used to train the model. The second is a validation set used to assess model performance.

```{r}
in.training <- createDataPartition(train.clean1$classe, p=0.70, list=F)
train.final <- train.clean1[in.training,]
validate.final <- train.clean1[-in.training,]
```


## Model Development

**1 - Train model**

The training data-set is used to fit a Random Forest model because it automatically selects important variables and is robust to correlated covariates & outliers in general. The Cross-validation technique assesses how the results of a statistical analysis will generalize to an independent data set. In 5-fold cross-validation, the original sample is randomly partitioned into 5 equal sized sub-samples. 

```{r}
control.parms <- trainControl(method="cv", 5)
rf.model <- train(classe ~ ., data=train.final, method="rf",
                 trControl=control.parms, ntree=251)
rf.model
```

**2 - Estimate performance**

The model fit using the training data is tested against the validation data. Predicted values for the validation data are then compared to the actual values. This allows forecasting the accuracy and overall out-of-sample error, which indicate how well the model will perform with other data.

```{r}
rf.predict <- predict(rf.model, validate.final)
confusionMatrix(validate.final$classe, rf.predict)
```

**3 - Results**
The accuracy of this model is `0.9940527` and the Overall Out-of-Sample error is `0.0059473`.

##Run the model

The model is applied to the test data to produce the results.

```{r}
results <- predict(rf.model, 
                   test.clean1[, -length(names(test.clean1))])
results
```

##Appendix - Decision Tree Visualization

```{r}
treeModel <- rpart(classe ~ ., data=train.final, method="class")
fancyRpartPlot(treeModel)
```


